Modeling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from biclaR

### 2024-05-03 paper/PaperCEUS/PaperCEUS_rev1.tex
solve bib

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% TODO: Currently lineno needs to be loaded after amsmath because of conflict
% https://github.com/latex-lineno/lineno/issues/5
\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can replace a high proportion of private car
trips (Martens, 2007). In The Netherlands, which has the highest mode
share of cycling in the world, cycling accounts for more than a third of
all trips to and from rail stations at the `home' end of the journey,
greatly increasing the ability of the transport system (Rietveld, 2000).
This approach to reducing car dependency and associated externalities
requires interventions and programs to make bicycling more appealing (La
Paix et al., 2021). The resulting public investments can have
significant social and environmental benefits (International Transport
Forum, 2017). Despite the benefits of cycling-PT intermodality, the
potential of this combination is often overlooked in transport planning
(La Paix et al., 2021).

The potential of cycling as a complementary mode of PT is substantial
worldwide, especially in cities with established public transport
networks or substantial ambitions to develop them. In the Lisbon
metropolitan area (LMA) the largest metropolitan area in Portugal, the
modal share of cycling is low, but the potential for cycling as a
complementary mode of PT is high. According to the latest mobility
survey conducted in 2018 (INE, 2018), the LMA registered a total of 5.3
million daily trips, with only 0.5\% by bicycle. Car modal share was
58.4\%, while PT accounted for 15.5\% (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in
the same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys.\\
53\% trips are up to 5 km distance, and 71\% up to 10 km. Nevertheless,
29\% of trips are longer than 10 km, which requires the use of motorized
modes, or active modes in combination with public transport.

\begin{figure}
\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
\end{figure}

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) (Presidência do
Conselho de Ministros, 2019), the Lisbon's Metropolitan Department of
Transport commissioned \emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the planning, design, and
development of a metropolitan cycling network (Félix et al., 2022).

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels (Lovelace et
al., 2017). The PCT has been extended to include cycling potential for
travel to school in England (Goodman et al., 2019) and other trip types
in other countries.\footnote{See \href{https://www.npt.scot}{npt.scot}
  and \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{modeling-origin-destination-trips}{%
\subsection{Modeling Origin-Destination
trips}\label{modeling-origin-destination-trips}}

The mobility survey data (INE, 2018) is the basis of the baseline
scenario and trip rates presented in this paper. Conducted in the
pre-pandemic period (2017), this OD dataset represents the most
comprehensive and up-to-date information on urban mobility in Portuguese
metropolitan areas (Lisbon and Porto).

We used `jittering' to disaggregate the OD data, resulting in a wide
spatial distribution of trip origins and destinations (Lovelace et al.,
2022b). The method works by sampling `sub-points' (nodes on the
transport network represented in OpenStreetMap in this case) and using
these instead of a single point (typically the centroid) to represent
trip origins and destinations for each zone. This method then
distributes the trips to desire lines connecting the subpoints based on
a `disaggregation threshold' which determines the maximum number of
trips that can be represented by a single desire line.

@@

{\centering \includegraphics[width=1\linewidth,]{img/jitter}

@@

\caption{Representation of desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
\end{figure}

@@

\hypertarget{modeling-routes}{%
\subsection{Modeling routes}\label{modeling-routes}}

@@

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} (Pereira
et al., 2021), which allows for great flexibility in configuring
estimated route types, and which proven to provide most accurate route
networks for the city of Lisbon (Lovelace et al., 2022a). \texttt{r5r}
can calculate multi-modal routes using PT combined with other modes. It
enables the identification of the most direct or safest cycling routes,
using the Level of Traffic Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively. Different routing profiles enable
decision-makers to plan for different bicycle user typologies and/or for
different city cycling maturity levels (Félix et al., 2017).

@@

\hypertarget{modeling-intermodality}{%
\subsection{Modeling intermodality}\label{modeling-intermodality}}

The intermodality scenario considers trips made by PT in which cycling
is used for the first and last legs. We restricted our analysis to the
first and last legs with a combined length of up to 5 km (for example: 1
km from origin to interface \emph{A} plus 4 km from interface \emph{B}
to destination) or up to 25 minutes on bike travel-time. This
conservative approach was adopted to capture the fact that cycling
stages as part of a multi-modal trip are likely to be shorter than
cycling-only trips (van Mil et al., 2021). Furthermore, we have imposed
restrictions on PT usage, limiting it to trips without PT transfers, and
within a duration of up to 2 hours (120 minutes). Additionally, we have
only included PT modes that can easily accommodate bicycles, such as
trains, ferries, trams, and inter-municipal bus lines equipped with bike
racks (Figure \ref{fig:map1}). These restrictions can be eased in the
future when testing more developed policy interventions to enhance
intermodality between cycling and PT, considering both the vehicle and
infrastructure perspectives.

@@

Figure \ref{fig:map2} illustrates the resulting bicycle routes to access
the main PT interfaces in the LMA.

@@

\caption{Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
\end{figure}

\hypertarget{assessing-socio-environmental-benefits}{%
\subsection{Assessing socio-environmental
benefits}\label{assessing-socio-environmental-benefits}}

For the \emph{cycling legs of the journey} (first and last legs),
socio-environmental impacts were estimated, using the Health Economic
Assessment Tool (HEAT) for Cycling v5.2 (Kahlmeier et al., 2017), from
the World Health Organization, and the
\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
package}\footnote{\texttt{HEATaaS} is under development. For more
  information contact
  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. The
use of this package made it possible to run multiple scenarios with few
changes in input values, making the interaction with HEAT more reliable
when reproducing runs.\\
The HEAT tool provided estimates on the shifting from car to cycling for
a short term time horizon (i.e., one year) and the long term (i.e., ten
years). It estimates the differences between two considered scenarios.
In this case: one baseline scenario, with data from the mobility survey,
and one cycling potential scenario in which targets of 4\% and 10\% of
cycling levels were achieved, transferred from car trips. We considered
two dimensions: \emph{social} --- including the physical activity, air
pollution exposure, and road casualties; and \emph{environmental} ---
including CO\textsubscript{2}eq emissions and other pollutants.

@@

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
methods and reference values (Ntziachristos \& Samaras, 2020) for a Tier
3 detail level. We used a family-size vehicle, EURO standard, and
gasoline or diesel fuel. All trips were considered to be made under
urban conditions and at an average speed of 15 km/h during rush hour
periods. Since the average distance traveled per trip influences the
overconsumption and emissions from cold-start engine operation, we
estimated energy and emission factors for different ranges of trips at
500-meter intervals.\\
An equation was then used to calculate emission factors for the two
types of fuel, for each type of pollutant, whose explanatory variables
are driving speed (\(speed\), in km/h) and average trip distance
(\(l_{trip}\), in km/trip). Thus, the emission factors
(\(EF_{fuel,l_{trip},speed}\), in g/km) can be calculated using equation
\ref{eq:fatoremissaoauto}.

@@

The used values consider that 64\% of the car fleet was diesel in
2022\footnote{See
  \href{https://smi.ine.pt/Indicador/Detalhes/10837?LANG=EN}{Statistics
  Portugal: Stock road vehicles statistic}.}. In addition, we assumed an
occupancy rate of 1.6 passengers \emph{per} car (INE, 2018). Finally,
the final emissions for each trip (\(E_{pollutant}\), in g/trip) are
derived from the equation \ref{eq:emissaoauto}.

@@

Regarding PT, we considered the emission factor values reported in the
environmental and sustainability reports of the PT operators in the LMA
(Carris, 2020; CP, 2020; Metropolitano de Lisboa, 2020; Transtejo,
2014). In particular, for the urban train and tram -- with 100\%
electric traction -- only CO\textsubscript{2}eq emissions were
considered (resulting from the production of electricity, considering a
``well-to-tank'' approach), since the other pollutants are not emitted
locally.

The conversion of avoided emissions into avoided welfare loss and
respective monetary valuation was based on the EU Guide to Cost-benefit
Analysis (Sartori et al., 2014) and the best up-to-date reference values
for the various gases (Bickel et al., 2006; Nash et al., 2003; Sartori
et al., 2014): 8.44 €/ton for CO, 2,867.85 €/ton for
NO\textsubscript{X}, 340,969.27 €/ton for PM\textsubscript{10}, 7,169.62
€/ton for VOC and 35.85 €/ton for CO\textsubscript{2}eq. The social
impacts are in avoided premature mortality. This result is finally
monetized using the \emph{Statistical Value of Life} for Portugal:
€3,055,358/fatality (Silva et al., 2021). We updated all the monetary
reference values of the literature based on the annual inflation rate in
Portugal for 2022\footnote{See
  \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
  Portugal tool for inflation rate estimates between years}.}, and our
10-years estimations assumed a discount rate of 5\% and inflation of
3\%. See \protect\hyperlink{research-data}{Research Data} for all the
input values we used.

\hypertarget{results-and-discussion}{%
\section{Results and Discussion}\label{results-and-discussion}}

@@

Shifting from car to cycling in combination with PT can reduce annual
CO\textsubscript{2}eq emissions by 11,500 to 28,500 tons per year. These
figures represent a 2.7\% reduction in Lisbon's transport emissions
(Lisboa E-Nova, 2023), a small but important component of wider
transport decarbonization measures. The 10-year socio-environmental
benefits account for €125 million to €325 million, depending on the
cycling targets.

The environmental impacts represent less than 2\% of the
socio-environmental benefits (in value) from replacing car trips to
bicycle in first-and-last legs. For the PT segment, we did not estimate
the social impacts from substituting car trips. One of the main
socio-environmetal benefits, valued after monetization, comes from the
increase in physical activity (Félix et al., 2023). Although there are
also social benefits form shifting car trips to PT, its health benefits
would not be as high as shifting to cycling. The literature shows that
the Metabolic Equivalent Tasks (MET) for ``riding in a bus or a train''
is 1.3 plus the ``walking for transportation'' as 3.5, while ``driving a
car'' is 2.5 (Ainsworth et al., 2011). The difference between these
activities - shifting from car to PT - is not very obvious when compared
to shifting from car to cycling, whose MET is about 6.8. Nevertheless,
future works should also encompass the estimation of the social impacts
for the PT leg of the journey, shifting from car.

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

@@

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

@@

\hypertarget{research-data}{%
\section*{Research data}\label{research-data}}
\addcontentsline{toc}{section}{Research data}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-MET2011}{}}%
Ainsworth, B. E., Haskell, W. L., Herrmann, S. D., Meckes, N., Bassett
Jr, D. R., Tudor-Locke, C., Greer, J. L., Vezina, J., Whitt-Glover, M.
C., \& Leon, A. S. (2011). {2011 Compendium of Physical Activities: A
second update of codes and MET values}. \emph{Medicine \& Science in
Sports \& Exercise}, \emph{43}(8), 1575--1581.
\url{https://doi.org/10.1249/MSS.0b013e31821ece12}

\leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
Bickel, P., Friedrich, R., Burgess, A., Fagiani, P., Hunt, A., Jong, G.
D., \& Laird, J. (2006). \emph{{HEATCO - Developing Harmonised European
Approaches for Transport Costing and Project Assessment. Deliverable 5,
Proposal for Harmonised Guidelines.}}
\url{https://trimis.ec.europa.eu/sites/default/files/project/documents/20130122_113653_88902_HEATCO_D5_summary.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
Carris. (2020). \emph{Relatório de sustentabilidade 2019 - demostração
não financeira}. {Carris - Companhia Carris de Ferro de Lisboa, E.M.,
S.A.} \url{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
CP. (2020). \emph{Relatório de sustentabilidade 2019}. {CP - Comboios de
Portugal, E.P.E.}
\url{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
Félix, R., Lovelace, R., \& Moura, F. (2022). \emph{{biclaR - Ferramenta
de apoio ao planeamento da rede ciclável na área metropolitana de
Lisboa}}. {CERIS - Instituto Superior Técnico and Transportes
Metropolitanos de Lisboa}. \url{https://biclar.tmlmobilidade.pt}

\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
Félix, R., Moura, F., \& Clifton, K. J. (2017). Typologies of urban
cyclists: Review of market segmentation methods for planning practice.
\emph{Transportation Research Record}, \emph{2662}(1), 125--133.
\url{https://doi.org/10.3141/2662-14}

\leavevmode\vadjust pre{\hypertarget{ref-Felix2023ES}{}}%
Félix, R., Orozco-Fontalvo, M., \& Moura, F. (2023). Socio-economic
assessment of shared e-scooters: Do the benefits overcome the
externalities? \emph{Transportation Research Part D: Transport and
Environment}, \emph{118}, 103714.
\url{https://doi.org/10.1016/j.trd.2023.103714}

\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
Goodman, A., Rojas, I. F., Woodcock, J., Aldred, R., Berkoff, N.,
Morgan, M., Abbas, A., \& Lovelace, R. (2019). Scenarios of cycling to
school in england, and associated health and carbon impacts: Application
of the {`}propensity to cycle tool{'}. \emph{Journal of Transport \&
Health}, \emph{12}, 263--278.
\url{https://doi.org/10.1016/j.jth.2019.01.008}

\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
INE. (2018). \emph{Mobilidade e funcionalidade do território nas {Áreas
Metropolitanas do Porto e de Lisboa}: 2017}. {Instituto National de
Estatística}.
\url{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt}

\leavevmode\vadjust pre{\hypertarget{ref-internationaltransportforum_integrating_2017}{}}%
International Transport Forum. (2017). \emph{Integrating {Urban Public
Transport Systems} and {Cycling}}. {Organisation for Economic
Co-operation and Development}.

\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
Kahlmeier, S., Götschi, T., Cavill, N., Castro Fernandez, A., Brand, C.,
Rojas Rueda, D., Woodcock, J., Kelly, P., Lieb, C., Oja, P., et al.
(2017). \emph{Health economic assessment tool ({HEAT}) for walking and
for cycling: Methods and user guide on physical activity, air pollution,
injuries and carbon impact assessments}.
\url{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-lapaix_role_2021}{}}%
La Paix, L., Cherchi, E., \& Geurs, K. (2021). Role of perception of
bicycle infrastructure on the choice of the bicycle as a train feeder
mode. \emph{International Journal of Sustainable Transportation},
\emph{15}(6), 486--499.
\url{https://doi.org/10.1080/15568318.2020.1765223}

\leavevmode\vadjust pre{\hypertarget{ref-LisboaENova}{}}%
Lisboa E-Nova. (2023). \emph{{Observatórios Lisboa - Infos Emissões
Gases com Efeito Estufa}}. {Agência de Energia e Ambiente de Lisboa}.
\url{https://observatorios-lisboa.pt/info_emissoes.html}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
Lovelace, R., Félix, R., \& Carlino, D. (2022a). Exploring jittering and
routing options for converting origin-destination data into route
networks: Towards accurate estimates of movement at the street level.
\emph{The International Archives of the Photogrammetry, Remote Sensing
and Spatial Information Sciences}, \emph{XLVIII-4/W1-2022}, 279--286.
\url{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
Lovelace, R., Félix, R., \& Carlino, D. (2022b). Jittering: A
computationally efficient method for generating realistic route networks
from origin-destination data. \emph{Findings}.
\url{https://doi.org/10.32866/001c.33873}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% TODO: Currently lineno needs to be loaded after amsmath because of conflict
% https://github.com/latex-lineno/lineno/issues/5
\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}

@@

@@

\section{Introduction}\label{introduction}

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can replace a high proportion of private car
trips \citep{MARTENS2007326}. In The Netherlands, which has the highest
mode share of cycling in the world, cycling accounts for more than a
third of all trips to and from rail stations at the `home' end of the
journey, greatly increasing the ability of the transport system
\citep{RIETVELD200071}. This approach to reducing car dependency and
associated externalities requires interventions and programs to make
bicycling more appealing \citep{lapaix_role_2021}. The resulting public
investments can have significant social and environmental benefits
\citep{internationaltransportforum_integrating_2017}. Despite the
benefits of cycling-PT intermodality, the potential of this combination
is often overlooked in transport planning \citep{lapaix_role_2021}.

The potential of cycling as a complementary mode of PT is substantial
worldwide, especially in cities with established public transport
networks or substantial ambitions to develop them. In the Lisbon
metropolitan area (LMA) the largest metropolitan area in Portugal, the
modal share of cycling is low, but the potential for cycling as a
complementary mode of PT is high. According to the latest mobility
survey conducted in 2018 \citep{IMOB}, the LMA registered a total of 5.3
million daily trips, with only 0.5\% by bicycle. Car modal share was
58.4\%, while PT accounted for 15.5\% (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in
the same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys.\\
53\% trips are up to 5 km distance, and 71\% up to 10 km. Nevertheless,
29\% of trips are longer than 10 km, which requires the use of motorized
modes, or active modes in combination with public transport.

\begin{figure}
\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
\end{figure}

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) \citep{ENMAC},
the Lisbon's Metropolitan Department of Transport commissioned
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the planning, design, and
development of a metropolitan cycling network \citep{felix2023}.

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels
\citep{lovelace2017}. The PCT has been extended to include cycling
potential for travel to school in England \citep{goodman2019} and other
trip types in other countries.\footnote{See
  \href{https://www.npt.scot}{npt.scot} and
  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

\section{Methods}\label{methods}

\subsection{Modeling Origin-Destination
trips}\label{modeling-origin-destination-trips}

The mobility survey data \citep{IMOB} is the basis of the baseline
scenario and trip rates presented in this paper. Conducted in the
pre-pandemic period (2017), this OD dataset represents the most
comprehensive and up-to-date information on urban mobility in Portuguese
metropolitan areas (Lisbon and Porto).

We used `jittering' to disaggregate the OD data, resulting in a wide
spatial distribution of trip origins and destinations
\citep{Lovelace2022Jittering}. The method works by sampling `sub-points'
(nodes on the transport network represented in OpenStreetMap in this
case) and using these instead of a single point (typically the centroid)
to represent trip origins and destinations for each zone. This method
then distributes the trips to desire lines connecting the subpoints
based on a `disaggregation threshold' which determines the maximum
number of trips that can be represented by a single desire line.

@@

{\centering \includegraphics[width=1\linewidth,]{img/jitter_new}

@@

\caption{Flow-weighted random sample of 10,000 desire lines in the Lisbon metropolitan area between districts of the 18 municipalities, without jittering (left) and with jittering (right).}\label{fig:jitter}
\end{figure}

@@

\subsection{Modeling routes}\label{modeling-routes}

@@

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package}
\citep{r5r}, which allows for great flexibility in configuring estimated
route types, and which proven to provide most accurate route networks
for the city of Lisbon \citep{Lovelace2022exploring}. \texttt{r5r} can
calculate multi-modal routes using PT combined with other modes. It
enables the identification of the most direct or safest cycling routes,
using the Level of Traffic Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively. Different routing profiles enable
decision-makers to plan for different bicycle user typologies and/or for
different city cycling maturity levels \citep{felix2017}.

@@

\subsection{Modeling intermodality}\label{modeling-intermodality}

The intermodality scenario considers trips made by PT in which cycling
is used for the first and last legs. We restricted our analysis to the
first and last legs with a combined length of up to 5 km (for example: 1
km from origin to interface \emph{A} plus 4 km from interface \emph{B}
to destination) or up to 25 minutes on bike travel-time. This
conservative approach was adopted to capture the fact that cycling
stages as part of a multi-modal trip are likely to be shorter than
cycling-only trips \citep{vanmil_insights_2021}. Furthermore, we have
imposed restrictions on PT usage, limiting it to trips without PT
transfers, and within a duration of up to 2 hours (120 minutes).
Additionally, we have only included PT modes that can easily accommodate
bicycles, such as trains, ferries, trams, and inter-municipal bus lines
equipped with bike racks (Figure \ref{fig:map1}). These restrictions can
be eased in the future when testing more developed policy interventions
to enhance intermodality between cycling and PT, considering both the
vehicle and infrastructure perspectives.

@@

Figure \ref{fig:map2} illustrates the resulting bicycle routes to access
the main PT interfaces in the LMA. It is noticeable the high potential
of the train interfaces to attract car-to-PT substituting trips.

@@

\caption{Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool). Larger the line width, higher the segment's car to bicycle shift potential. Darker the line color, lower the quietness level.}\label{fig:map2}
\end{figure}

\subsection{Assessing socio-environmental
benefits}\label{assessing-socio-environmental-benefits}

For the \emph{cycling legs of the journey} (first and last legs),
socio-environmental impacts were estimated, using the Health Economic
Assessment Tool (HEAT) for Cycling v5.2 \citep{HEAT}, from the World
Health Organization, and the
\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
package}\footnote{\texttt{HEATaaS} is under development. For more
  information contact
  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. The
use of this package made it possible to run multiple scenarios with few
changes in input values, making the interaction with HEAT more reliable
when reproducing runs.\\
The HEAT tool provided estimates on the shifting from car to cycling for
a short term time horizon (i.e., one year) and the long term (i.e., ten
years). It estimates the differences between two considered scenarios.
In this case: one baseline scenario, with data from the mobility survey,
and one cycling potential scenario in which targets of 4\% and 10\% of
cycling levels were achieved, transferred from car trips. We considered
two dimensions: \emph{social} --- including the physical activity, air
pollution exposure, and road casualties; and \emph{environmental} ---
including CO\textsubscript{2}eq emissions and other pollutants.

@@

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
methods and reference values \citep{COPERT} for a Tier 3 detail level.
We used a family-size vehicle, EURO standard, and gasoline or diesel
fuel. All trips were considered to be made under urban conditions and at
an average speed of 15 km/h during rush hour periods. Since the average
distance traveled per trip influences the overconsumption and emissions
from cold-start engine operation, we estimated energy and emission
factors for different ranges of trips at 500-meter intervals.\\
An equation was then used to calculate emission factors for the two
types of fuel, for each type of pollutant, whose explanatory variables
are driving speed (\(speed\), in km/h) and average trip distance
(\(l_{trip}\), in km/trip). Thus, the emission factors
(\(EF_{fuel,l_{trip},speed}\), in g/km) can be calculated using equation
\ref{eq:fatoremissaoauto}.

@@

The used values consider that 64\% of the car fleet was diesel in
2022\footnote{See
  \href{https://smi.ine.pt/Indicador/Detalhes/10837?LANG=EN}{Statistics
  Portugal: Stock road vehicles statistic}.}. In addition, we assumed an
occupancy rate of 1.6 passengers \emph{per} car \citep{IMOB}. Finally,
the final emissions for each trip (\(E_{pollutant}\), in g/trip) are
derived from the equation \ref{eq:emissaoauto}.

@@

Regarding PT, we considered the emission factor values reported in the
environmental and sustainability reports of the PT operators in the LMA
\citep{Carris2019s, Metro2019s, CP2019s, Transtejo2014}. In particular,
for the urban train and tram -- with 100\% electric traction -- only
CO\textsubscript{2}eq emissions were considered (resulting from the
production of electricity, considering a ``well-to-tank'' approach),
since the other pollutants are not emitted locally.

The conversion of avoided emissions into avoided welfare loss and
respective monetary valuation was based on the EU Guide to Cost-benefit
Analysis \citep{EuropeanCommission2014} and the best up-to-date
reference values for the various gases
\citep{EuropeanCommission2014, bickel2006, UNITE}: 8.44 €/ton for CO,
2,867.85 €/ton for NO\textsubscript{X}, 340,969.27 €/ton for
PM\textsubscript{10}, 7,169.62 €/ton for VOC and 35.85 €/ton for
CO\textsubscript{2}eq. The social impacts are in avoided premature
mortality. This result is finally monetized using the \emph{Statistical
Value of Life} for Portugal: €3,055,358/fatality \citep{ANSR2021}. We
updated all the monetary reference values of the literature based on the
annual inflation rate in Portugal for 2022\footnote{See
  \href{https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ipc}{Statistics
  Portugal tool for inflation rate estimates between years}.}, and our
10-years estimations assumed a discount rate of 5\% and inflation of
3\%. See \hyperref[research-data]{Research Data} for all the input
values we used.

\section{Results and Discussion}\label{results-and-discussion}

@@

Shifting from car to cycling in combination with PT can reduce annual
CO\textsubscript{2}eq emissions by 11,500 to 28,500 tons per year. These
figures represent a 2.7\% reduction in Lisbon's transport emissions
\citep{LisboaENova}, a small but important component of wider transport
decarbonization measures. The 10-year socio-environmental benefits
account for €125 million to €325 million, depending on the cycling
targets.

The environmental impacts represent less than 2\% of the
socio-environmental benefits (in value) from replacing car trips to
bicycle in first-and-last legs. For the PT segment, we did not estimate
the social impacts from substituting car trips. One of the main
socio-environmetal benefits, valued after monetization, comes from the
increase in physical activity \citep{Felix2023ES}. Although there are
also social benefits form shifting car trips to PT, its health benefits
would not be as high as shifting to cycling. The literature shows that
the Metabolic Equivalent Tasks (MET) for ``riding in a bus or a train''
is 1.3 plus the ``walking for transportation'' as 3.5, while ``driving a
car'' is 2.5 \citep{MET2011}. The difference between these activities -
shifting from car to PT - is not very obvious when compared to shifting
from car to cycling, whose MET is about 6.8. Nevertheless, future works
should also encompass the estimation of the social impacts for the PT
leg of the journey, shifting from car.

@@

\section{Conclusion}\label{conclusion}

@@

\section*{Acknowledgements}\label{acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

@@

\section*{Research data}\label{research-data}
\addcontentsline{toc}{section}{Research data}

@@

\renewcommand\refname{References}
\bibliography{bibliography.bib}
